Automatic discovery of cell types and microcircuitry from neural connectomics

### 2014-06-23 docs/arxiv.tex
incorporated yarden's first set of suggestions

%\documentclass{nature}
\documentclass{article}
\usepackage[english,american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage[disable]{todonotes}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
%\usepackage{algorithm}
%\usepackage{algorithmic}
\usepackage{amsmath}
%\usepackage{subcaption}
%\usepackage{multicol}
\usepackage{csquotes}
\usepackage{pdfpages}
\usepackage{placeins}

@@

\title{Automatic discovery of circuit structure from neural connectomics}
\author{Eric Jonas$^1$, Konrad Kording$^{2, 3,4}$}

@@

\begin{small}
\begin{enumerate}
  \item Department of Brain and Cognitive Sciences, Massachusetts Institute of Technology, Cambridge, Massachusetts
  \item Department of Physical Medicine and Rehabilitation, Northwestern University and Rehabilitation Institute of Chicago, Chicago, Illinois
  \item Department of Physiology, Northwestern University, Chicago, Illinois
  \item Department of Applied Mathematics, Northwestern University, Chicago, Illinois
\end{enumerate}
\end{small}

\begin{abstract}
  New techniques produce massive data about neural connectivity,
  necessitating new analysis methods to discover underlying
  structure. It has long been assumed that discovering the local
  patterns of microcircuitry is the key to understanding neural
  function. Here we developed a nonparametric Bayesian technique that
  identifies neuron types and circuitry patterns in connectomics
  data. We show that the approach recovers known neuron types in the
  retina, reveals interesting structure in the nervous system of
  c. elegans, and automatically discovers the structure of
  microprocessors. Our approach extracts structural meaning from
  connectomics, enabling new approaches of deriving anatomical
  insights from these emerging datasets.

\end{abstract}

\section*{Introduction}
Emerging techniques \autocite{Morgan2013,Zador2012} promise to
quantify the location of each neuron within a volume of interest,
along with its connections to all other neurons. Far exceeding the
capacity of neuroanatomists to trace small circuits, this leads to
enormous datasets that quantify aspects of the nervous system. The impact of
this kind of high-throughput transition has been seen before -- 
rise of sequencing techniques necessitated the
development of novel computational methods to understand genomic
structure, ushering in an era of bioinformatics as an independent
discipline \autocite{Koboldt2013}.

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined local connectivity patterns
has been prominent in many areas, ranging from sensory (e.g. retina,
\autocite{Masland2001} via processing (e.g. neocortex \autocite{Mountcastle1997})
to movement (e.g. spinal cord) \autocite{Grillner2005}. Indeed, such
repeated patterns also exists in human-made computing circuits. It
remains an important challenge to develop algorithms to use anatomical
data, e.g. connectomics, to back out underlying connectivity motifs.

The discovery of structure is a crucial aspect of network
science. Early approaches focused on global graph properties, such as
the types of scaling present in the network \autocite
{WattsStrogatz1998} .  While this approach leads to an understanding
of global network, more recent work aims at identifying very small-scale
repeat patterns, or “motifs” in networks\autocite{Milo2002}

The discovery of structure in probabilistic graphs is a crucial aspect
in machine learning. Commonly used algorithms include
community-based-detection methods \autocite{Girvan2002}, and
stochastic block models \autocite{Nowicki2001}.  While these
approaches can incorporate the probabilistic nature of neural
connections \autocite{Hill2012} they do not generally consider the
known fact that the probability of connections between neurons depends
on distance. Combining such probabilistic models with a spatial
component to model distance promises to improve the detection of
microcircuitry from neural data.

When it comes to neuroscience and other computing systems, we expect
patterns of connectivity much more complex than traditional motifs,
exhibiting a strong spatial dependence arising from the complex
genetic, chemical, and activity-based neural development process.

Here we describe a Bayesian non-parametric model that can discover
circuit structure automatically from connectomics data: both the cell
types and their spatial patterns of interconnection. We apply it to
three computing systems: the mouse retina connectome
\autocite{Helmstaedter2013}, the c. elegans connectome
\autocite{White1986}, and a ``connectome'' of a classical
microprocessor \autocite{James2010}. Note we do not explicitly
consider cellular morphology, absolute position (cell layer), or any
chemical or genetic information. Yet in all cases, we discover
cell-type to cell-type connectivity that depends on spatial distance
and these connectivity functions depend on the two involved cell
types. Comparing the cell types discovered by the algorithms with
those obtained by anatomists based on orthogonal information reveals a
high degree of agreement. We present a scalable approach to infer
microcircuitry from connectivity data.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{overview.ai}}
  \caption{An overview of our method. a. Input data is
a connectivity matrix indicating whether cell $i$ (row)
synapses on cell $j$ (column). The location of each cell in space is known. 
b. Our algorithm discovers hidden cell types in this connectivity data
by assuming all cells of a type share a distance-dependent connectivity 
profile, c.), with cells of other types. This results in a clustering 
of the cells by those hidden types. b.) shows the cell connectivity
matrix with cells of the same type grouped together. d.) shows
the learned probability of connection between our different types
at various distances, illustrating short-range and long-range
connectivity patterns.}
\label{fig:overview}
\end{figure}

\section{Results}
We use a Bayesian structure discovery approach that is based on
prior knowledge about the way positions and cell types affect the
probability of connections between cells, a so-called generative
model. We term our model the ``infinite spatial-relational model'' (iSRM), 
in keeping with the existing nonparametric literature.  
The probability of connections between pairs of neurons
decreases with distance, but how high the baseline probability of
connection is and how quickly it decays with distance depends on the
two cell-types. Note that this simplistic prior information is all that 
we tell our model; no additional physiological information is included.

We combine annealing with Markov Chain Monte Carlo to
perform simultaneously joint posterior inference over the structure,
parameters, and hyperparameters of the model (see Methods). This
procedure automatically learns the number of cell types, which cells 
belong to which types, along with the 
the distance and type dependent connectivity probabilities (see
fig~\ref{fig:overview}). These results can then be compared with the
results of previous anatomical studies that have used other aspects of
neurons, such as chemical properties and cell-morphology to divide
neurons into types.

Before we can use a probabilistic model to analyze real data we first
need to assess if the model works properly. We thus started out by
simulating data for which we know the correct structure and comparing
the estimated structure based on the algorithm (see methods) with the
one we used for simulation. We find that the model does a good job of
recovering the structure we used to generate the data
(fig~\ref{fig:synthetic}a) Indeed, it does far better than a simpler
and well established model (the infinite stochastic block model, or
iSBM) that assumes that there is no distance dependence to the
connectivity. The model converges relatively quickly to an estimate of
the most probable values for the cell types, which is enabled by using
a combination of simulated annealing and parallelized Markov chain
Monte Carlo (see methods for details). It appears that the model
behaves the way it should and allows application to relatively large
datasets.

\begin{figure}
  \centering 
  \centerline{\includegraphics{synthetic.ai}}
  \caption{Correct recovery of true numbers of hidden classes
in synthetic data. a.) Other models over-cluster the number
of classes as they fail to take distance into account, whereas
our model finds close to the true number of classes. b.) ARI, a measure
of ``correct clustering'', for
different true class counts, between our model and a traditional
iSBM. c.) For synthetic data, 
the iSBM overclusters, finding a large number of spatially-localized types. 
In contrast, the infinite spatial-relational model recovers the true spatial
extent of the underlying types.}
\label{fig:synthetic}
\end{figure}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics{mouseretina.ai}}
  \caption{Discovering cell classes in the mouse retina connectome. 
a.) Input connectivity data for 950 cells for which soma positions were known. b. clustered connectivity matrix. c.) connectivity diagram showing our clusters, as
well as the cell depth and known cell type of each cell. Our model agrees
with the cell types traditionally identified by anatomists, and identifies 
laminar-specific connectivity patterns. d.) Connectivity between our
clusters as a function of distance -- the cluster consisting primarily of
retinal ganglion cells (lower left-center) exhibits the expected near and
far connectivity. e. The spatial distribution of our cell types -- each cell type tesselates space.}
\label{fig:mouseretina}
\end{figure}

The retina of the mouse \autocite{Masland2001} is an example where we
should expect to have connectivity patterns that are well approximated
by our generative model. It is known that there are multiple classes
of cells that can be grouped into ganglion cells that transmit
information to the rest of the brain, bipolar cells that connect
between different cells, and amacrine cells that feed into the
ganglion cells \autocite{cite some review}. Recent research
\autocite{Helmstaedter2013} has produced a large dataset containing
both the types of cells from orthogonal approaches, and also the
connectivity matrix between all reconstructed cells. We want to know
to which level we can classify neurons into types based on this
connectivity only.

The algorithm took 8 hours to converge to a locally optimal solution,
dividing neurons into a set of cell types (fig~\ref{fig:mouseretina}c, each
wedge is a type). For each pair of neurons there is a specific
distance dependent connection probability (fig~\ref{fig:mouseretina}b,c,d),
which is well approximated by the model fit. Moreover, each type of
cell is rather isotropically distributed across space
(fig~\ref{fig:mouseretina}e) as should be expected for true cell types. The
algorithm is thus able to extract meaningful structure from the data.

Comparing the results of the algorithm to other information sources
allows evaluating the quality of the type determination. Our types
closely reflect the (anatomist-determined) segmentation of cells into
retinal ganglion, narrow amacrine, medium/wide amacrine, and bipolar
cells (fig~\ref{fig:mouseretina}c, outermost ring). We find that the
types we find tend to reflect the known laminar distribution in the
retina (fig~\ref{fig:mouseretina}c, middle ring), which is exciting as
the algorithm did not actually use this type of information.

@@

\todo{Do we assign any of the unknown cells correctly?  are types 72-79
“different” or simply unknown?}

We compared the results of our model to the traditional iSBM to see if
the addition of spatial information was necessary. We find that, as
expected from our synthetic experiments, the iSBM massively
over-segmented the data, finding spatially-localized “communities” of
cells which bore little relation to known types.  The average spatial
extent of the types found by the iSBM was $14.6\pm8.1 \mu m$. However,
in the ground truth by anatomists the average distance was $34.3\pm7.6
\mu m$ and $35.3\pm 2.4 \mu m$, quite close to the $34.5\pm 1.0 \mu m$
discovered by our algorithm. Modeling the distance-dependence prevents
over-segmentation and thus allows more meaningful estimates.

@@

The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons deterministically known. This
incredible cross-organism consistency led to early mapping of the
c. elegans chemical and electrical (gap junction) connectome. Unlike
the retina, only the motor neurons in c. elegans exhibit regular
distribution in space, being distributed along the body axis. Most
interneurons are concentrated in various ganglia that project
throughout the entire animal, and the sensory neurons are primarily
located in a small number of anterior ganglia.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=5in]{celegans.ai}}
  \caption{Discovering cell types in the \textit{c. elegans}
    connectome. a.) Initial cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportional to synapse count between
    those cells. b.) Clustered adjacency matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). }
  \label{fig:celegans}
\end{figure}

Using both the chemical and electrical connectivity (see methods), we
determined the underlying clusters explained by connectivity and
distance (fig~\ref{fig:celegans}a). A superficial inspection of the results shows
clustering into groups consisting roughly homogeneously of motor
neurons, sensory neurons, and interneurons. Closer examination reveals
agreement with the classifications originally outlined by Brenner in
1986.

We identify cell types that reflect the known motor/non-motor neuron
classes, even though this system lacks the strong repeat
microcircuitry our model was designed for.  Motor-neuron types AS, DA,
and VA, all exclusively postsynaptic, are clustered together, as are
motor-neuron types VD and DD. Traditional types VC, DB, and VB also
mostly share a cluster. Various head motor neurons, including types
SMD, RMD, and RIM are clustered together. Interneurons with known
anatomically-distinct connectivity patterns, such as RIA (2 cells) or
AVA (2 cells) are clustered into pure types. The algorithm even
correctly places the single-cell type DVB by itself. Note our
clustering does not perfectly reflect known divisions -- several
combinations of head and sensory neurons are combined, and a
difficult-to-explain group of 2 VD, 2 VB, a VA, and a DD neuron are
grouped together. Our identified cell types thus reflect a
``coarsening'' of known types, based entirely on connectivity and
distance information, even when the organism exhibits substantially
less spatial regularity than the retina.

\subsection{Microprocessor}
\begin{figure}
  \centering 
  \centerline{\includegraphics{mos6502.ai}}
  \caption{Discovering transistor types in the MOS 6502
    microprocessor. a.) The clusters and connectivity versus distance
    for connections between Gate and C1, Gate and C2, and C1 and C2
    terminals on a transistor. Red and teal types have a terminal
    pulled down to ground and mostly function as inverters. Purple
    class are clocked, stateful transistors, orange control the ALU
    and yellow control the special data bus. b.) The spatial
    distribution of the transistors in each cluster. }
\label{fig:mos6502}
\end{figure}

To show the applicability of our method to other connectome-style
datasets, we obtained the spatial location and interconnectivity of
the transistors in a classic microprocessor, the MOS Technology 6502
(used in the Apple II) \autocite{James2010}. Computer architects use
common patterns of transistors when designing circuits, with each
transistor having a ``type'' in the circuit. We identified a region of
the processor with complex but known structure containing the primary
8-bit registers X, Y, and S (fig~\ref{fig:mos6502}).

Our algorithm identifies areas of spatial homogeneity that mirror the
known structure in the underlying architectural circuit, segmenting
transistor types recognizable to computer architects. Using the
original schematics, we see that one identified type contains the
``clocked'' transistors, which retain digital state. Two other types
contain transistors with pins C1 or C2 connected to ground, mostly
serving as inverters.  An additional identified type controls the
behavior of the three registers of interest (X, Y, and S) with respect
to the SB data bus, either allowing them to latch or drive data from
the bus. The repeat patterns of spatial connectivity are visible in
figure~\ref{fig:mos6502}a, showing the man-made horizontal and
vertical layout of the same types of transistors.

\section{Discussion}
We have presented a machine learning technique that allows cell types
and microcircuitry to be discovered using only distance and
connectivity data. We have shown its applicability to a newly
published connectomics dataset from the mouse retina, the classical
connectivity of the worm c. elegans, and the connectivity of a
historical man-made microprocessor. We have found that distance and
connectivity alone are sufficient to discover cell types and
microcircuitry that were known to exist in the systems based on decades
of previous research which had utilized extensive additional
information.

For probabilistic models like ours, no known solution exists to
exactly find the most probable parsing of the neurons into cell-types
and connectivity patterns . We employ a collection of Markov-chain
Monte carlo techniques (see Methods) but while different
initializations converge to similar ultimate values, we can never
realistically obtain the global optimum. There are a broad range of
techniques that may offer better approximations to the global optimum
\autocite{Wanga2012} and future work could adapt them to
the problem we addressed here.

Our inference becomes slower as the amount of data increases. Our
algorithm required 30 hours for 1000 neurons and could take months for
substantially longer datasets. Scaling this class of probabilistic
model is an active area of research, and recent results in both
variational methods \autocite{Hoffman2013} and spectral learning
  \autocite{Anandkumar2012} promise substantial speed increases.

  Larger datasets will allow algorithms to distinguish more distinct
  types \todo{Add supplemental data} and we expect closer agreement
  with existing anotomical knowledge as more data become available.
  Moreover, in general, for such problems precision increases with the
  size of the dataset and the cells that we have are not sufficient to
  statistically distinguish all the cell types known in anatomy (such
  as the $\sim 70$ in the retina). Still, using only connectivity and
  distance it is possible to meaningfully divide neurons into types.

@@

There exist a range of previous approaches to the discovery of
neural microcircuitry \todo{Do we want a citation here?}. These
generally involve a great deal of manual labor and ad-hoc
determination of what constitutes a “type” of cell -- to this day
there are disagreements in the literature as to the “true” types in
the mammalian retina. Much as phylogenomics has changed our
understanding of animal ontologies, modern large scale data will allow
the efficient unbiased discovery of cell types and circuits. The sheer
amount of available data demands the introduction of algorithmic
approaches.

>>>>

%\documentclass{nature}
\documentclass{article}
\usepackage[american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage{todonotes}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
%\usepackage{algorithm}
%\usepackage{algorithmic}
\usepackage{amsmath}
%\usepackage{subcaption}
%\usepackage{multicol}
\usepackage{csquotes}
\usepackage{pdfpages}
\usepackage{placeins}

@@

\title{Automatic discovery of cell types and microcircuitry from neural connectomics}
\author{Eric Jonas$^1$, Konrad Kording$^{2, 3,4}$}

@@

\begin{small}
\begin{enumerate}
  \item Electrical Engineering and Computer Science, University of California, Berkeley
  \item Department of Physical Medicine and Rehabilitation,
    Northwestern University and Rehabilitation Institute of Chicago,
    Chicago, Illinois
  \item Department of Physiology, Northwestern University, Chicago, Illinois
  \item Department of Applied Mathematics, Northwestern University, Chicago, Illinois
\end{enumerate}
\end{small}

\begin{abstract}
  Neural connectomics has begun producing massive amounts of data,
  necessitating new analysis methods to discover the biological and
  computational structure. It has long been assumed that discovering
  neuron types and their relation to microcircuitry is crucial to
  understanding neural function. Here we developed a nonparametric
  Bayesian technique that identifies neuron types and microcircuitry
  patterns in connectomics data. It combines the information
  traditionally used by biologists, including connectivity, cell body
  location and the spatial distribution of synapses.  We show that the
  approach recovers known neuron types in the retina and enables
  predictions of connectivity, better than simpler algorithms. It also
  can reveal interesting structure in the nervous system of
  c. elegans, and automatically discovers the structure of a
  microprocessor.  Our approach extracts structural meaning from
  connectomics, enabling new approaches of automatically deriving
  anatomical insights from these emerging datasets.
\end{abstract}

\section*{Introduction}
Emerging connectomics techniques \autocite{Morgan2013,Zador2012}
promise to quantify the location and connectivity of each neuron
within a tissue volume. These massive datasets will far exceed the
capacity of neuroanatomists to manually trace small circuits, thus
necessitating computational, quantitative, and automatic methods for
understanding neural circuit structure.  The impact of this kind of
high-throughput transition has been seen before -- rise of sequencing
techniques necessitated the development of novel computational methods
to understand genomic structure, ushering in  bioinformatics
as an independent discipline \autocite{Koboldt2013}.

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined, type-dependent local
connectivity patterns (microcircuits) has been prominent in many
areas, from sensory (e.g. retina, \autocite{Masland2001} to processing
(e.g. neocortex \autocite{Mountcastle1997}) to movement (e.g. spinal
cord) \autocite{Grillner2005}. These sorts of repeated computing
patterns are a common feature of computing systems, even arising in
human-made computing circuits. It remains an important challenge to
develop algorithms to use anatomical data, e.g. connectomics, to
automatically back out underlying microcircuitry.

The discovery of structure is a crucial aspect of network
science. Early approaches focused on global graph properties, such as
the types of scaling present in the network \autocite
{WattsStrogatz1998}.  While this approach leads to an understanding
of the global network, more recent work aims at identifying very small-scale
repeat patterns, or “motifs” in networks\autocite{Milo2002}. These motifs
are defined not between different node types, but rather represent repeated
patterns of topology.

The discovery of structure in probabilistic graphs is a crucial aspect
in machine learning. Commonly used algorithms include
community-based-detection methods \autocite{Girvan2002}, and
stochastic block models \autocite{Nowicki2001}.  While these
approaches can incorporate the probabilistic nature of neural
connections \autocite{Hill2012} they do not incorporate the additional
richer structure present in connectomics data -- the location of cell
bodies, the spatial distribution of synapses, and the distances
between neurons. Of particular importance is that the probability of
connections has a strong spatial component, a factor that is hard to
reconcile with many other methods. A model attempting to fully capture
the variation in the nervous system should take into account the broad
set of available features.

When it comes to neuroscience and other computing systems, we expect
patterns of connectivity much more complex than traditional motifs,
exhibiting a strong spatial dependence arising from the complex
genetic, chemical, and activity-based neural development processes.

To address these challenges, here we describe a Bayesian
nonparametric model that can discover circuit structure automatically
from connectomics data: the cell types, their spatial patterns of
interconnection, and the locations of somata and synapses. We show
that by incorporating this additional information, our model both
accurately predicts the connection as well as agrees
with human neuroanatomists as to the identification of cell types.

We primarily focus on the recently-released mouse retina connectome
\autocite{Helmstaedter2013}, but additionally examine the c. elegans
connectome \autocite{White1986}, and then ``connectome'' of a
classical microprocessor \autocite{James2010}. Comparing the cell
types discovered by the algorithms with those obtained manually by
human anatomists reveals a high degree of agreement. We present a
scalable approach to infer microcircuitry from connectomics data.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{fig1b.ai}}
  \caption{An overview of our method. a. Input data is a connectivity
    matrix indicating whether cell $i$ (row) synapses on cell $j$
    (column), a distance matrix between cell soma, b. the depth of the
    cell soma, and the c. depth profile of the cells synaptic
    connectivity. d. Our algorithm discovers hidden cell types in this
    connectivity data by assuming all cells of a type share a
    distance-dependent connectivity profile, similar depth, and a
    similar synpatic density profile, with cells of other types.  This
    results in a clustering of the cells by those hidden
    types. d. shows the cell connectivity matrix with cells of the
    same type grouped together. e.) shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. f. shows a connectivity profile with virtually no
    probability of connectivity. We can plot this these types g.) to
    show the relationship and spatial connections, and how probable
    various types are to connect to one another. h. The connectivity
    between our discovered types changes as a function of distance. }

\label{fig:overview}
\end{figure}

\section{Results}

We start with the generic assumption that cell types exist -- a cell
is a member of a single, unobserved type -- and these types affect
soma depth, distribution of synapses, as well as a cell type and
distance dependent connection probability. From this assumption we
develop a generative Bayesian model that estimates the underlying cell
types and how they connect.  We start with a model for connectivity,
the infinite stochastic block model (iSBM)\autocite{Kemp2006a,Xu2006},
which has been shown to meaningfully cluster connection graphs while
learning the number of hidden groups, or types. We extend this
approach by adding distance dependence to model salient aspects of
microcircuitry. We additionally model cell body depth as a mixture of
Gaussians and each cell’s synapse density profile as mixture of
mixtures of Gaussians.

For example, retinal ganglion cells may synapse on nearby, but not far
away, amacrine cells, with ganglion cells being superficial and having
a broad distribution of synapses.  This procedure automatically learns
the number of cell types, which cells belong to which types, along
with the distance and type dependent connectivity probabilities (see
fig~\ref{fig:overview}). We apply our algorithm to datasets from mouse
retina, c-elegans and a historical microprocessor. Anatomists classify
cells based on many features, giving us a meaningful baseline to
compare against.

Before we can use a probabilistic model to analyze real data we first
need to assess if the model works properly. We thus simulate data for
which we know the correct structure and comparing the estimated
structure based on the algorithm (see methods) with the one we used
for simulation. We find that the model does a good job of recovering
the correct number of cell types, (fig~\ref{fig:synthetic}a), the the
cell identities (fig~\ref{fig:synthetic}b), and the spatial extent of
each type (fig~\ref{fig:synthetic}c).  Indeed, it does far better than
the existing infinite stochastic block model that assumes that there
is no distance dependence to the connectivity. The model converges
relatively quickly to an estimate of the most probable values for the
cell types, which is enabled by using a combination of simulated
annealing and parallelized Markov chain Monte Carlo (see methods for
details). It appears that the model behaves the way it should and
allows application to relatively large datasets.

\begin{figure}
  \centering 
  \centerline{\includegraphics{synthetic.ai}}
  \caption{Correct recovery of true numbers of hidden types in
    synthetic data when incorporating spatial information. a.) The
    infinite stochastic block model (which only uses connectivity
    information) over-estimates the number of classes as it fails to
    take distance into account, whereas our modeling of the
    combination of distance and connectivity finds close to the true
    number of classes. b.) ARI, a measure of ``correct clustering'',
    for different true class counts, between our model and a
    traditional iSBM. c.) For synthetic data, type discovery based
    only on connectivity are tiny spatially-localized groups -- in
    contrast, our model recovers the true spatial extent of the
    underlying types.  }
\label{fig:synthetic}
\end{figure}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics{mouseretina.ai}}
  \caption{Discovering cell classes in the mouse retina connectome.
    a.) Input connectivity data for 950 cells for which soma positions
    were known. b. clustered connectivity matrix. c.) connectivity
    diagram showing our clusters, as well as the cell depth and
    anatomist-labeled cell type of each cell. Our model shows moderate
    agreement with the cell types traditionally identified by
    anatomist laminar-specific connectivity patterns. 
    d.) Connectivity between our clusters as a function
    of distance -- the cluster consisting primarily of retinal
    ganglion cells (lower left-center) exhibits the expected near and
    far connectivity. e.)  The spatial distribution of our cell types
    -- each cell type tesselates space.}
\label{fig:mouseretina}
\end{figure}

The mouse retina \autocite{Masland2001} is a neural circuit which we
expect to have connectivity patterns that are well approximated by our
generative model. It is known that there are multiple classes of cells
that can be broadly grouped into: ganglion cells that transmit
information to the rest of the brain, bipolar cells that connect
between different cells, and amacrine cells that feed into the
ganglion cells. Recent research \autocite{Helmstaedter2013} has
produced a large dataset containing both the types of cells from
orthogonal approaches, and also the connectivity matrix between all
reconstructed cells.

The algorithm took less than 2 hours to perform inference, dividing
neurons into a set of cell types (fig~\ref{fig:mouseretina}c, each
wedge is a type). For each pair of neurons there is a specific
distance dependent connection probability
(fig~\ref{fig:mouseretina}b,c,d), which is well approximated by the
model fit. Moreover, each type of cell is rather isotropically
distributed across space (fig~\ref{fig:mouseretina}e) as should be
expected for true cell types.

Comparing the results of the algorithm to other information sources
allows evaluating the quality of the type determination. Our types
closely reflect the (anatomist-determined) segmentation of cells into
retinal ganglion, narrow amacrine, medium/wide amacrine, and bipolar
cells (fig~\ref{fig:mouseretina}c, outermost ring). We find that the
types we find tend to reflect the known laminar distribution in the
retina (fig~\ref{fig:mouseretina}c, middle ring).

@@

Our fully Bayesian model produces a distribution over probable
clusterings.  Figure~\ref{fig:mouseretina_compare} shows this
posterior distribution as a cell-cell coassignment matrix, sorted to
find maximum block structure. Each large, dark block represents a
collection of cells believed with strong probability to be of the same
type. When we plot (fig~\ref{fig:mouseretina_compare}b) the
anatomist-derived cell types along the left, we can see that each
block consists of a roughly-homogeneous collection of types.

We evaluate our model along three sets of parameters
(Fig.\ref{fig:mouseretina_compare}): how closely does our clustering agree
with neuroanatomists' knowledge?  Given two cells, how accurately can
our model predict the link between them? And how closely does the
spatial extent (within a layer) of our identified types agree with the
known neuroanatomists.

As a first measure we compare link prediction accuracy across the
methods (Fig.\ref{fig:mouseretina_compare} B AUC, red).  We find that
given the dataset many techniques allow for good link-predictive
accuracy. All the methods allow decent link prediction with an AUC in
the .9 range. However, our algorithm clearly outperforms the simple
statistical models that only use connectivity.

As a second measure we compare link prediction accuracy across the
methods (Fig.\ref{fig:mouseretina_compare} B ARI, blue). We find that our
algorithm far outperforms the controls. We also find that when it is
based on more of the same information used by anatomists use then it
gets better at agreeing with these anatomists. In particular, using
connectivity, distance, synapse distribution and soma depth leads to
the highest ARI. When using the available information the algorithm
produces a good fit to human anatomist judgments.

Finally we look at the spatial extent of the discovered types both
within a layer and between layers (Fig.\ref{fig:mouseretina_compare}
C). We see that, in the absence of distance information, mere
connectivity information results in types which only span a small
region of space -- essentially “local cliques”. Incorporation of
distance information results in types which span the entire extent of
the layer. The depth variance of all models continues to be
substantially larger than that predicted by human anatomists -- future
directions of work include attempting to more strongly encode this
prior belief of laminarity.

\begin{figure}
  \centering 
  \centerline{\includegraphics{comparison.ai}}
  \caption{a. Cell-cell assignment matrix reflecting the posterior
    distribution of clusterings found by our model. Large contiguous
    black areas are cells that were considered the same type in the
    majority of samples. b. The cells and their true types sorted to
    agree with the coassignment matrix. c. A comparison of the
    predictive accuracy (Area under the curve) for hand-labeled
    anatomical data, versus inclusion of additional sources of
    information. d. ) The spatial extent (in depth and area) of the
    types identified by humans and our various algorithmic
    approaches.}

\label{fig:mouseretina_compare}
\end{figure}

@@

The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons developmentally determined,
leading to early measurement of the connectome. Unlike the retina,
only the motor neurons in c. elegans exhibit regular distribution in
space -- along the body axis. Most interneurons are
concentrated in various ganglia that project throughout the entire
animal, and the sensory neurons are primarily located in a small
number of anterior ganglia.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=7in]{otherdata.ai}}
  \caption{Discovering connectivity and type in both
    \textit{c. elegans} and the MOS 6502 microprocessor.  a.) Initial
    \textit{c. elegans} cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportional to synapse count between
    those cells. b.) Clustered adjacency matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). d.) The clusters and connectivity versus
    distance for connections between Gate and C1, Gate and C2, and C1
    and C2 terminals on a transistor. Red and teal types have a
    terminal pulled down to ground and mostly function as
    inverters. Purple class are clocked, stateful transistors, orange
    control the ALU and yellow control the special data bus. e.) The
    spatial distribution of the transistors in each cluster.}
  \label{fig:otherdata}
\end{figure}

Using both the chemical and electrical connectivity (see methods), we
determined the underlying clusters explained by connectivity and
distance (fig~\ref{fig:otherdata}a). A superficial inspection of the results shows
clustering into groups consisting roughly homogeneously of motor
neurons, sensory neurons, and interneurons. Closer examination reveals
agreement with the classifications originally outlined by Brenner in
1986.

We identify cell types that reflect the known motor/non-motor neuron
classes, even though this system lacks the strong repeat
microcircuitry our model was designed for.  Motor-neuron types AS, DA,
and VA, all exclusively postsynaptic, are identified as a common type,
as are motor-neuron types VD and DD. Traditional types VC, DB, and VB
also mostly share a cluster. Various head motor neurons, including
types SMD and RMD, are clustered together. Interneurons with known
anatomically-distinct connectivity patterns, such as AVA (2 cells),
are clustered into pure types.The algorithm even correctly places the
single-cell types DVB and DVC by themselves.

Note our clustering does not perfectly reflect known divisions --
several combinations of head and sensory neurons are combined, and a
difficult-to-explain group of mostly VB and DB motor neuron types,
with VC split between various groups. Our identified cell types thus
reflect a ``coarsening'' of known types, based entirely on
connectivity and distance information, even when the organism exhibits
substantially less spatial regularity than the retina.

\subsection{Microprocessor}

To show the applicability of our method to other connectome-style
datasets, we obtained the spatial location and interconnectivity of
the transistors in a classic microprocessor, the MOS Technology 6502
(used in the Apple II) \autocite{James2010}. Computer architects use
common patterns of transistors when designing circuits, with each
transistor having a ``type'' in the circuit. We identified a region of
the processor with complex but known structure containing the primary
8-bit registers X, Y, and S (fig~\ref{fig:otherdata}).

Our algorithm identifies areas of spatial homogeneity that mirror the
known structure in the underlying architectural circuit, segmenting
transistor types recognizable to computer architects. Using the
original schematics, we see that one identified type contains the
``clocked'' transistors, which retain digital state. Two other types
contain transistors with pins C1 or C2 connected to ground, mostly
serving as inverters.  An additional identified type controls the
behavior of the three registers of interest (X, Y, and S) with respect
to the SB data bus, either allowing them to latch or drive data from
the bus. The repeat patterns of spatial connectivity are visible in
figure~\ref{fig:otherdata}a, showing the man-made horizontal and
vertical layout of the same types of transistors.

\section{Discussion}
We have presented a machine learning technique that allows cell types
and microcircuitry to be discovered from connectomics data.  We have
shown its applicability to a newly published connectomics dataset from
the mouse retina. When compared to existing methods, we show how the
incorporation of all of this data yields results that combine both
high link-prediction accuracy and high agreement with human
anatomists.  I also reveals interesting structure when applied to the
classical connectivity of the worm c. elegans, and the connectivity of
a historical man-made microprocessor. We have found that combining the
available data types allows us to discover cell types and
microcircuitry that were known to exist in the systems based on
decades of previous research and allows good prediction of
connectivity.

For our probabilistic models, no known solution exists to
exactly find the most probable parsing of the neurons into cell-types
and connectivity patterns. We employ a collection of Markov-chain
Monte carlo techniques (see Methods) but while different
initializations converge to similar ultimate values, we can never
realistically obtain the global optimum. There are a broad range of
techniques that may offer good approximations to the global optimum
and future work could adapt them to
find more precise solutions to our problem.

For our probabilistic model, inference becomes slower as the amount of
data increases. Our algorithm required several hours for 1000
neurons. Scaling this class of probabilistic model is an active area
of research, and recent results in both variational methods
\autocite{Hoffman2013} and spectral learning \autocite{Anandkumar2012}
and future work could adapte them to find faster approximate solutions
to our problem.

Larger datasets will allow algorithms to distinguish more distinct
types and we expect closer agreement with existing anatomical
knowledge as more data become available.  Moreover, in general, for
such problems precision increases with the size of the dataset and the
cells that we have are not sufficient to statistically distinguish all
the cell types known in anatomy (such as the $\sim 70$ in the
retina). Still, using only connectivity and distance it is possible to
meaningfully divide neurons into types.

@@

There exist a range of previous approaches to the discovery of neural
microcircuitry\autocite{Mountcastle1957, Douglas1991, Bartho2004,
  Freund1998}.  These generally involve a great deal of manual labor
and ad-hoc determination of what constitutes a “type” of cell -- to
this day there are disagreements in the literature as to the “true”
types in the mammalian retina. Much as phylogenomics has changed our
understanding of animal ontologies, modern large scale data will allow
the efficient unbiased discovery of cell types and circuits. The sheer
amount of available data demands the introduction of algorithmic
approaches.
